Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-08-26 manuscript/main.tex
update figure 2 legend

As several prior studies have demonstrated associations between specific gut taxa
and atypical neurocognition and neurological disorders
\cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021,
      magnussonRelationshipsDietrelatedChanges2015,mayneris-perxachsMicrobiotaAlterationsProline2022,
      needhamGutderivedMetaboliteAlters2022},
we sought to determine if specific taxa or gene functions were associated
with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism were associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we separately considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age).
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was significantly associated with variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0124, q \textless{} 0.001),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0119, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants (Figure 1G).
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and  gene functions
annotated with UniRef90s clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles, both generated from metagenomic sequences, being tightly coupled
(Figure 1H, p \textless{} 0.001). Other functional
groupings (Enzyme commission level4 - ECs
\cite{bairochENZYMEDatabase20002000},
and KEGG orthologues - KOs
\cite{kanehisaKEGGResourceDeciphering2004})
overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (15.9\%, p \textless{} 0.01) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression models (LMs)
\cite{mallickMultivariableAssociationDiscovery2021}
for the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N\textsubscript{species} = 92 for 0--120 months, 
N\textsubscript{species} = 46 for 0--6 months, N\textsubscript{species} = 97 for 18--120 months).
Only \emph{Blautia wexlerae} was significantly associated (q value = 0.14, $\beta$ = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B).
By contrast, in children over 18 months of age,
several mirobial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae},
\emph{Asaccharobacter celatus} and \emph{Adelcreutzia equolifaciens}
(also known as \textit{Adlercreutzia equolifaciens} subsp. \textit{celatus}
and \textit{Adlercreutzia equolifaciens} subsp. \textit{equolifaciens} respectively
\cite{maruoAdlercreutziaEquolifaciensGen2008}),
and SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure 2B). \textit{Ruminococcus gnavus} was the only microbial species
that was significantly negatively associated with cognitive function score (q value = 0.18).

\begin{sidewaysfigure}[!htb]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure2.png}
    \caption{
        Taxa and gene functional groups are associated with cognitive function.
        (\textbf{A}) Volcano plot of multivariable linear model results showing the
        relationship between individual taxa and cognitive function score in
        children over 18 months of age, controlling for age, maternal education,
        and sequencing depth. Taxa that were significant after FDR correction (q
        \textless{} 0.2) are colored red. (\textbf{B}) For taxa that were significantly
        associated with cognitive function, heatmaps of prevalence, mean
        relative abundance, and correlation with cognitive function in different
        age groups. (\textbf{C-F}) Enrichment plots for selected neuroactive gene sets
        used in feature set enrichment analysis (FSEA). (\textbf{G}) summary of FSEA
        results across all samples (left) as well as in the under 6 months
        (middle) and over 18 months (right) subsets, colored based on the
        significance of the association. Markers indicate the individual
        correlation of genes within a gene set, and vertical bars represent the
        median correlation of that gene set.
    }
    \label{fig:2}
\end{sidewaysfigure}

>>>>

As several prior studies have demonstrated associations between specific gut taxa
and atypical neurocognition and neurological disorders
\cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021,
      magnussonRelationshipsDietrelatedChanges2015,mayneris-perxachsMicrobiotaAlterationsProline2022,
      needhamGutderivedMetaboliteAlters2022},
we sought to determine if specific taxa or gene functions were associated
with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism were associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we separately considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age).
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was significantly associated with variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0099, q \textless{} 0.01),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0119, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants (Figure 1G).
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0207, and  gene functions
annotated with UniRef90s clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0263, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging profiles (R\textsuperscript{2} = 0.258, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles, both generated from metagenomic sequences, being tightly coupled
(Figure 1H, p \textless{} 0.001). Other functional
groupings (Enzyme commission level4 - ECs
\cite{bairochENZYMEDatabase20002000},
and KEGG orthologues - KOs
\cite{kanehisaKEGGResourceDeciphering2004})
overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (taxa - 20.1\%, p \textless{} 0.01;
gene functions - 15.9\%, p \textless{} 0.05) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression models (LMs)
\cite{mallickMultivariableAssociationDiscovery2021}
for the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N\textsubscript{species} = 116 for 0--120 months, 
N\textsubscript{species} = 54 for 0--6 months, N\textsubscript{species} = 136 for 18--120 months).
No species were significantly associated (q \textless{} 0.20) with
cognitive function in children under 6 months old after adjusting for
age and maternal education.
By contrast, in children over 18 months of age,
several mirobial species were significantly enriched
in children with higher cognitive function scores (Figure 2B),
including \emph{Alistipes obesi},
\emph{Asaccharobacter celatus} (also known as \textit{Adlercreutzia equolifaciens} subsp. \textit{celatus}
\cite{maruoAdlercreutziaEquolifaciensGen2008}),
and SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure 2B). \textit{Sutterella wadsworthensis} was the only microbial species
that was significantly negatively associated with cognitive function score (q value = 0.165).

\begin{sidewaysfigure}[!htb]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure2.png}
    \caption{
        Taxa and gene functional groups are associated with cognitive function.
        (\textbf{A}) Volcano plot of multivariable linear model results showing the
        relationship between individual taxa and cognitive function score in
        children over 18 months of age (left) or across all ages (right),
        controlling for age, maternal education, and sequencing depth.
        Taxa that were significant after FDR correction (q
        \textless{} 0.2) are colored red. (\textbf{B}) For taxa that were significantly
        associated with cognitive function,
        heatmaps of correlation with cognitive function in different age groups.
        (\textbf{C}) Multivariate linear models showing the relationship
        between individual taxa and domain-specific skill assessments
        from the MSEL for children under 6m (left), 18m to 48m (right),
        or over across all ages (right)
        controlling for age, maternal education, and sequencing depth.
        Asterisks in (A) and (C) indicate significance (q \textless{} 0.2)
        after multiple hypothesis correction.
        (\textbf{D-E}) Enrichment plots for proprionate degradation I gene set
        used in feature set enrichment analysis (FSEA) in children under 6m (D)
        or over 18m (E). (\textbf{G}) summary of FSEA
        results in children under 6m of age (left)
        over 18 months (middle) or across all ages (right), colored based on the
        significance of the association. Markers indicate the individual
        correlation of genes within a gene set,
        and vertical bars colored bars represent the
        median correlation of that gene set.
        Vertical dotted lines within each panel represent the median T statistic
        for all genes in that age group.
    }
    \label{fig:2}
\end{sidewaysfigure}
